---
title: "Multivariate Statistical Analysis Problem set 3"
author: Edoardo Bandoni  - Giorgia Cerase  -Lorenzo Sala - Stefano Sperti  -  Francesco Virgili
date: "Università di Torino"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{=tex}
\textit{Old Faithful is a hydrothermal geyser in Yellowstone National Park in the state of Wyoming, U.S.A., and
is a popular tourist attraction. Its name stems from the supposed regularity of its eruptions. The data set
faithful (package dataset) comprises 272 observations, each of which represents a single eruption and
contains two variables corresponding to eruptions, the duration in minutes of the eruption, and waiting,
the time until the next eruption, also in minutes. Figure below shows a plot of the time to the next eruption
versus the duration of the eruptions. It can be seen that the time to the next eruption varies considerably,
although knowledge of the duration of the current eruption allows it to be predicted more accurately.}
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datasets)
library(mclust)
library(ellipse)
library(TargetScore)
data <- as.data.frame(scale(faithful))
```

```{r}
head(faithful)
```

# Exercise 1

\textit{Obtain a 2-clusters solution via kmeans. Plot the observations with colors determined by cluster memberships as in panel (i) of Figure 9.1 of Bishop (2006).}

```{r}
km.data <- kmeans(data,center=2)
centroids <- km.data$centers
labels <- km.data$cluster
col.index <- c("blue","red")[labels]
plot(data,pch=16,col=col.index,cex=1.3,xlim=c(-2.2,2.2),ylim=c(-2.2,2.2))
points(centroids,cex=2.5,col="white",pch=4,lwd=7)
points(centroids,cex=2.5,col=c("blue","red"),pch=4,lwd=3)
```

This code utilizes the K-means clustering algorithm to group
\texttt{faithful} points into two distinct clusters. The \texttt{kmeans}
function performs the clustering, assigning each data point to a cluster
based on its similarity to the cluster's centroid (central point).
K-means algorithm starts by randomly choosing k centroids (cluster
centers) within the data space. It then assigns each data point to the
closest centroid based on a distance measure (Euclidean distance). Once
all points are assigned, the centroids are recalculated as the mean of
their assigned points. These last two steps are repeated until the
centroids no longer significantly change, indicating convergence.

# Exercise 2

```{=tex}
\textit{Consider model-based clustering with Gaussian mixtures. Use mclust:: Mclust to fit a mixture with
two components, allowing for different covariance matrices. Reproduce the plot of panel (f) of Figure
9.8 of Bishop (2006), by adding contours of constant probability density for each component. Pay
attention to the color-coding of the observations obtained as combination of blue and red inks with
proportions equal to posterior probabilities.}
```
```{r}
model <- Mclust(data,G=2)
mean1 <- as.numeric(model$parameters$mean[,1])
mean2 <- as.numeric(model$parameters$mean[,2])
cov1 <- model$parameters$variance$sigma[,,1]
cov2 <- model$parameters$variance$sigma[,,2]
prob <- model$z
col.index2 <- rgb(red=prob[,1],green=0,blue=prob[,2],maxColorValue = 1)
plot(data,pch=16,cex=1.3,col=col.index2,xlim=c(-2.2,2.2),ylim=c(-2.2,2.2))
lines(ellipse(centre=mean1,cov1,level=0.68),lwd=8,col="white")
lines(ellipse(centre=mean1,cov1,level=0.68),lwd=4,col="red")
lines(ellipse(centre=mean2,cov2,level=0.68),lwd=8,col="white")
lines(ellipse(centre=mean2,cov2,level=0.68),lwd=4,col="blue")
```

In Figure 9.8 of Bishop (2006) represent a mixture of two Gaussians. The
centers of these Gaussians are initialized using the same values as
those for the K-means algorithm, and the precision matrices are
initialized to be proportional to the unit matrix. Specifically, Bishop
(2006) plot the result of the EM algorithm when have converged.

For the replication we utilized \texttt{Mclust}, which is a model-based
clustering method based on parameterized finite Gaussian mixture models.
Models are estimated using the EM algorithm initialized by hierarchical
model-based agglomerative clustering. The correct model is selected by
calculating the Bayesian Information Criterion (BIC). We specified the
parameter \texttt{G = 2} to denote the number of mixture components
(clusters) for which the BIC is calculated.

The Expectation-Maximization algorithm works as follows: In the Expectation Step, the probabilities that each data point belongs to each Gaussian component, based on the current parameter estimates (the responsibilities) are calculated. In the Maximization Step, the parameters of the Gaussian components (means, covariances, and mixing proportions) are updated using the responsibilities calculated in the previous step. This repeated process continues until convergence. In particular some controll are added such that the log likelihood increases with each iteration.



In the plot, we included the following command:
\texttt{col.index2 <- rgb(red = prob[, 1], green = 0, blue = prob[, 2], maxColorValue = 1)}:
This line creates a vector of colors based on the posterior
probabilities extracted earlier. It assigns a color to each data point
based on its posterior probability of belonging to the first and second
components. The \texttt{rgb} function is used to specify the colors
based on the red, green, and blue components, with the red component
determined by \texttt{prob[, 1]} and the blue component determined by
\texttt{prob[, 2]}, while the green component is fixed at 0. Our
scatterplot reveals that only two central data points deviate from the
dominant red and blue colors (likely due to their distance from the
centroids). The remaining data points likely belong to the correct
clusters with high probability, given the bimodal behavior exhibited by
the dataset.

We also plotted the 0.68 countour plot (one standard deviation for
Gaussian) for the two clusters.

# Exercise 3

```{=tex}
\textit{Many models of interest involve distributions which are not
tractable from a computational point of view, in particular in Bayesian
inference this is often the case for posterior distributions.}
```
Variational inference is an approximation schema used to deal with such
cases, with similarities with Monte Carlo methods, but with the
difference that variational inference is a deterministic approach.
Variational methods are based on the idea of optimizing a functional,
which is basically a function of functions; often, to do this we need to
restrict the range of functions on which the optimization is performed,
which leads to an approximate result.

Variational methods are particularly useful in the context of Bayesian
models, in which we have the observed variables denoted by $X$, the
parameters (which are also random variables) denoted by $\theta$, and
latent variables denoted by $U$, for simplicity we denote
$Z=(U,\theta)$; we wish to find a simpler distribution $q(Z)$, that we
choose in a restricted family of distributions, and that approximates
the posterior distribution $p(Z|X)$ which is usually intractable. The
main idea for is to find the distribution $q(Z)$ that minimizes the
Kullback-Leibler divergence with $p(Z|X)$, which is a measure of the
difference between two distribution and it is defined as
$$\text{KL}(q\,||\,p) = - \int q(Z) \ln \left( \frac{p(Z|X)}{q(Z)} \right) \,dZ$$
We notice that KL divergence it is not a proper distance but it is
always non-negative. Then we make use of the following relation:
$$\ln p(X) = L(q) + \text{KL}(q\,||\,p)$$ where
$$L(q) = \int q(Z) \ln \left( \frac{p(X, Z)}{q(Z)} \right) \, dZ$$ which
is easily proved 
\begin{align*}
    L(q) + \text{KL}(q\,||\,p)=&\int q(Z) \left(\ln \left( \frac{p(X, Z)}{q(Z)} \right)-\ln \left( \frac{p(Z|X)}{q(Z)}\right)\right) \, dZ\\
    =&\int q(Z)\ln \left( p(X)\right) \, dZ\\
    =&\ln p(X)
\end{align*} Thus, since the KL divergence is always non-negative,
minimizing it with respect to $q$ is equivalent to maximize $L$ with
respect to $q$, with the advantage that to maximize $L$ we do not need
to deal with $p(Z|X)$ which is intractable. We therefore assume some
restriction on $q$ to ensures tractability while maintaining flexibility
to approximate the true posterior well. In general models the only
restriction that we impose is that $$q(Z) = \prod_{i=1}^{N} q_i(Z_i).
$$ In particular, in our model we are dealing with a Gaussian mixture
model and we can assume a weaker restriction. We will restrict the
maximization problem only to the distributions $q$ which are such that
the distribution of the parameters and the one of the latent variables
factorize, that is $$q(Z)=q(\theta)q(U)$$ (recall that $U$ are the
latent variables and $\theta$ are the parameters that in our case
$\theta = \{\mu,\pi,\Lambda\}$). In particular we deal with Gaussian
mixture model. In particular, let define $U = \{ u_1, \ldots, u_N \}$ be
a $1 \times K$ binary vector with elements $u_{nk}$ for
$k=1, \ldots, K$. To define our model we define the joint distribution
of all random variables: \begin{equation}
  p(X, Z, \pi, \mu, \Lambda) = p(X \mid Z, \mu, \Lambda) p(Z \mid \pi) p(\pi) p(\mu , \Lambda)
\end{equation}
 $$ p(Z|\pi) = \prod_{n=1}^N\prod_{k=1}^K \pi_k^{u_{nk}} $$
conditional distribution of $Z$ given mixing coefficients. Since we are
a Gaussian model we define: $p(X | U, \mu, \Lambda)$ that is conditional
distribution of $X$ given $Z$, $\mu =\{ \mu_k\}$ and
$\Lambda =\{ \Lambda_k\}$.
$$ p(X | U, \mu, \Lambda) = \prod_{n=1}^N\prod_{k=1}^K \mathcal{N}(x_n|\mu_k,\Lambda_k^{-1})^{u_{nk}} $$.Then,
let define $p(\pi)$ the distribution of mixing coefficients where we
choose a Dirichlet distribution over the mixing coefficients since the
analysis is considerably simplified if we use conjugate prior
distributions and at the end we introduce $p(\mu, \Lambda)$, a
Gaussian-Wishart prior that governs the mean and precision of each
component.

Therefore, In our model the maximization can be made with respect to
only two factors separately. It can be proved that the solution to the
problem are given by

$$q(U)=\frac{1}{K_1}\exp\left\{E_{\theta}[\ln p(X,U,\theta)]\right\}$$
$$q(\theta)=\frac{1}{K_2}\exp\left\{E_{U}[\ln p(X,U,\theta)]\right\}$$
where $K_1,K_2$ are the normalization constants. (These equations are derived from 10.43 Bishop (2006)). Since one solution depends on the other, an
iterative algorithm is used to solve the problem, so we introduce the
variational EM algorithm. We will explain it in the particular case of
usage for a mixture of Gaussian model, in which the latent variable $U$
defines the affiliation to one between $K$ different Gaussian components
(clusters), and the parameters are:

```{=tex}
\begin{itemize}
\item $\mu=(\mu_1,...,\mu_K)$ the vector of means of the components;
\item $\Lambda=(\Lambda_1,...,\Lambda_K)$ the importance matrices of the components (the inverse of the covariances);
\item $\pi=(\pi_1,...,\pi_K)$ which are the mixing coefficients of the components (prior probabilities).
\end{itemize}
```
In particular, as previous define we have that $\Lambda$ is Wishart
distributed with parameters $W_0,\nu_0$ and that
$\mu_k|\Lambda_k\sim\mathcal{N}(m_0,(\beta_0\Lambda_k)^{-1})$ and $\pi$
has a Dirichelet distribution with parameter $\alpha_0$.

In this context we analyze the \texttt{R} function that we used to
reproduce image 10.6, which is an implementation of the variational EM
on a model like the one defined above. The function is called
\texttt{vbgmm} and it is from the \texttt{TargetScore} package
(documentation at <https://rdrr.io/bioc/TargetScore/src/R/vbgmm.R>, the function is also attached as appendix of the homework); its
main part consists in the initialization of the posterior probabilities
of the latent variable, which are stored in the matrix
$R=(r_{nk})_{N\times K}$, called responsibility matrix; then we have a
cycle of variational maximizations (\texttt{vmax}) and variational
expectations (\texttt{vexp}) steps, that finishes when one between the
convergence or the maximum number of iteration is reached.

The function \texttt{initialization}, basically chooses $K$ points in
the dataset as means of the Gaussian components and then assigns each
data point to a specific component, based on some measure of the
distance from each mean. Therefore, the initialization is random, hence
we cannot reproduce exactly the initial setting and the various
iterations in figure 10.6: as we'll see later the only panel well
reproduced is the one in which the convergence is attained.

The function \texttt{vmax} performs the variational maximization step,
in which the parameters (given in input) are updated according to the
following formulas: 
\begin{align*}
\alpha_k=\alpha_0+N_k\\
\beta_k &= N_k +\beta_0 \\
m_k &= \frac{1}{\beta_k}(\beta_0m_0+N_k\bar{x}_k) \\
W_k^{-1} &= W_0^{-1} + N_kS_k + \frac{\beta_0 N_k}{\beta_0+N_k}(\bar{x}_k - m_0)(\bar{x}_k - m_0)^T\\
\nu_k &= \nu_0 + N_k
\end{align*}
This are useful for update $q(\mu_k,\Lambda_k)$ that as previously
define is a Gaussian-Wishart distibution and is given by
$$q(\mu_k,\Lambda_k)=\mathcal{N}(\mu_k|m_k,(\beta_k\Lambda_k)^{-1})\mathcal{W}(\Lambda_k|W_k,\nu_k)$$
In particular, $N_k$ is the column sum of the responsibility matrix and
represents the numerosity of the k-th component, and
$$\bar{x}_k=\frac{1}{N_k}\sum_{n=1}^Nr_{nk}x_n$$
$$S_k=\frac{1}{N_k}\sum_{n=1}^Nr_{nk}(x_n-\bar{x}_k)(x_n-\bar{x}_k)^T$$
are the maximum likelihood estimate, and they are basically a weighted
sum of the ML estimates of the single components, weighted with the
posterior probabilities $r_{nk}$. The computation of the latter is
performed in the variational expectation step; we have that
$$E[z_{nk}]=r_{nk}$$ in particular it can be proven that
$$r_{nk}=\frac{1}{C}\rho_{nk}=\frac{1}{C}\exp\left\{E[\ln\pi_k]+\frac{1}{2}E[\ln|\Lambda_k|]-\frac{D}{2}\ln(2\pi)-\frac{1}{2}E_{\mu_k,\Lambda_k}[(x_n-\mu_k)^T\Lambda_k(x_n-\mu_k)]\right\}$$
where $C$ is the normalization constant. Notice that the computation of
these expectations depend on the parameters and their distributions, so
we see the dependence between one step and the other.

So in order to reproduce the last panel of figure 10.6, in which the
algorithm converged, we run the function with some fixed parameters:

```{r}
set.seed(220202)
n <- dim(data)[1]
p <- dim(data)[2]
k <- 6
alpha <- 0.0015
par <- list(alpha = rep(alpha,k),
              kappa = 1,
              m = colMeans(data),
              v = p+1,
              M = diag(1,p,p))
model <- vbgmm(data,init=k,prior=par)
```

Then we extrapolate the means of the components from the model output,
and compute the sample covariances $S_k$ (according to 10.53), which
we'll use to plot the ellipses:

```{r}
mk <- model$mu#cluster means
Nk <- colSums(model$full.model$R)#cluster cardinality
covk <- array(dim=c(2,2,6))
R <- model$R
for (i in 1:k) {
  covk[,,i] <- t(R[,i]*as.matrix(data-mk[,i]))%*%as.matrix(data-mk[,i])/Nk[i]
}
```

Finally we compute the expected mixing coefficients which will give us
the transparency of the color of the ellipses:

```{r}
pik <- (model$full.model$alpha)/(k*alpha+n)#mixing coefficients
```

Our model converged in 94 steps, so we present four phases, respectively
2,25,50,94 iterations:

```{r}
for (j in c(2,25,50,100)) {
  set.seed(220202)
  model <- vbgmm(data,init=k,prior=par,maxiter = j)
  mk <- model$mu#cluster means
  Nk <- colSums(model$full.model$R)#cluster cardinality
  pik <- (model$full.model$alpha)/(k*alpha+n)
  covk <- array(dim=c(2,2,6))
  R <- model$R
  for (i in 1:k) {
    covk[,,i] <- t(R[,i]*as.matrix(data-mk[,i]))%*%as.matrix(data-mk[,i])/Nk[i]
  }
  plot(data,col="green",cex=1,xlim=c(-3,3),ylim=c(-3,3))
  text(-2.5,2.5,labels=j)
  for (i in 1:k) {
    if (pik[i]>10^{-5}) {
      points(x=mk[1,i],y=mk[2,i],pch=18,cex=1.5,col=2)
      polygon(ellipse(centre=mk[,i],covk[,,i],level=0.68,),lwd=4,col=rgb(1,0,0,pik[i]),border="red")
    }
  }
}

```

As anticipated, the only figure perfectly resembling figure 10.6 is the
last one, in which the algorithm converged; the other are different
because the algorithm has a random initialization, and moreover we do
not know the exact parameters with which it was initialized. However, we
see that it proceeds in a similar fashion, starting with six different
components and then moving to three and ending up with two. Notice that
the components are actually always 6, but we plot them only when their
mixing coefficient is distinguishable from 0, as suggested by the book.

We tried also with different parametrization of the prior distibution $alpha_0 = 0.002$ , $m_0 = 0$, $beta_0 = 5$, $M =diag(6,p,p)$. 

```{r}
set.seed(220202)
n <- dim(data)[1]
p <- dim(data)[2]
k <- 6
alpha <- 0.002
par <- list(alpha = rep(alpha,k),
              kappa = 5,
              m = rep(0,p),
              v = p+1,
              M = diag(6,p,p))
model <- vbgmm(data,init=k,prior=par)
mk <- model$mu#cluster means
Nk <- colSums(model$full.model$R)#cluster cardinality
covk <- array(dim=c(2,2,6))
R <- model$R
for (i in 1:k) {
  covk[,,i] <- t(R[,i]*as.matrix(data-mk[,i]))%*%as.matrix(data-mk[,i])/Nk[i]
}
pik <- (model$full.model$alpha)/(k*alpha+n)#mixing coefficients
  plot(data,col="green",cex=1,xlim=c(-3,3),ylim=c(-3,3))
  for (i in 1:k) {
    if (pik[i]>10^{-5}) {
      points(x=mk[1,i],y=mk[2,i],pch=18,cex=1.5,col=2)
      polygon(ellipse(centre=mk[,i],covk[,,i],level=0.68,),lwd=4,col=rgb(1,0,0,pik[i]),border="red")
    }
  }
```
We can see that convergence occurs in nearly half the steps. This is because selecting the right hyperparameters aids in the model's convergence.





# Exercise 4

\textit{Reproduce Figure 10.7 of Bishop (2006) about determining the number of components by using the variational lower bound. Code should be provided in a separate file. Make a comparison with the best model selected by BIC method with mixtures up to 6 components (use **mclust::mclustBIC**)}

In a Bayesian Gaussian mixture model, the goal is to approximate the
true posterior distribution of the latent variables and model parameters
using a simpler distribution. The variational lower bound (ELBO) is used
as an objective function to optimize during the inference process. The
ELBO is defined as $$
\mathcal{L}= \mathbb{E}_q[log(p(X,Z,\pi,\mu,\Lambda))]-\mathbb{E}_q[log(q(Z,\pi,\mu,\Lambda))]
$$

```{r, message=FALSE, warning=FALSE, results='hide'}
Nobs <- 100
data <- as.data.frame(scale(faithful))
n <- dim(data)[1]
p <- dim(data)[2]
Lower <- matrix(0, ncol = 6,nrow=Nobs)

for (k in 2:6) {
  for (j in 1:Nobs) {
    alpha <- 0.0015
    par <- list(alpha = rep(alpha,k),
                kappa = 1,
                m = colMeans(data),
                v = p+1,
                M = diag(1,p,p))
    model <- vbgmm(data,init=k,prior=par,maxiter=120)
    Lower[j,k] <- max(model$L)
  }
}

plot(jitter(rep(2,Nobs)),Lower[,2], ylim = c(-1.615,-1.603), xlim = c(2,6), pch=3,
     col="red", xlab="Number of Components", ylab="Variational Lower Bound")
points(jitter(rep(3,Nobs)),Lower[,3], pch=3, col="red")
points(jitter(rep(4,Nobs)),Lower[,4], pch=3, col="red")
points(jitter(rep(5,Nobs)),Lower[,5], pch=3, col="red")
points(jitter(rep(6,Nobs)),Lower[,6], pch=3, col="red")
```

Figure 10.7 of Bishop (2006) represents the plot of the variational
lower bound versus the number of components in the Gaussian mixture
model, for the Old Faithful data, showing a peak at $K=2$ components.
For each value of $K$, the model is trained from $100$ different random
starts and the result, shown as '$+$' symbols plotted with small random
horizontal perturbations so that they can be distinguished. Note that
the solution for $K=2$ find suboptimal local maxima, but that this
happens infrequently.

To replicate the results, we adapt the code from Exercise 3 and compute
the Variational Bayesian Gaussian Mixture Model (VB-GMM) multiple times.
The \texttt{`vbgmm(data, init=k, prior=par, maxiter=120)} function
returns a data frame describing the VB-GMM, which includes the variable
\texttt{L} representing the Evidence Lower Bound (ELBO). For each value
of the number of clusters \texttt{K} we compute the variational lower
bound $100$ times and store the values in a matrix called
\texttt{Lower}. At this point, we plot the values of the ELBO,
specifying the '$+$' symbols by \texttt{pch=3} and we use
\texttt{jitter()} to create small random horizontal perturbations so
that the values are distinguishable.

It is noteworthy to observe that the peak in $K=2$ has a precise
meaning, i.e. that it is the number of clusters that maximizes the log
likelihood and hence the model is well approximated by a Variational
Bayesian Gaussian Mixture Model with two clusters.

```{r}
model <- mclustBIC(data,G=seq(from=1,to=6,by=1))
model
plot(model)
```

When using the \texttt{mclustBIC} function on the dataset with the range
of clusters specified as $1$ to $6$, the Bayesian Information Criterion
(BIC) indicates that the top three models, ranked in order, are as
follows: EEE with $3$ clusters, VEV with $2$ clusters, and VEE with $3$
clusters.

In particular \texttt{mclustBIC} provide the lower bound of the EM
algorithm for normal mixture models with a variety of covariance
structures, and functions for simulation from these models.

```{=tex}
\begin{itemize}
  \item \textbf{EII:} equal volume, round shape (spherical covariance)
  \item \textbf{VII:} varying volume, round shape (spherical covariance)
  \item \textbf{EEI:} equal volume, equal shape, axis parallel orientation (diagonal covariance)
  \item \textbf{VEI:} varying volume, equal shape, axis parallel orientation (diagonal covariance)
  \item \textbf{EVI:} equal volume, varying shape, axis parallel orientation (diagonal covariance)
  \item \textbf{VVI:} varying volume, varying shape, equal orientation (diagonal covariance)
  \item \textbf{EEE:} equal volume, equal shape, equal orientation (ellipsoidal covariance)
  \item \textbf{EEV:} equal volume, equal shape, varying orientation (ellipsoidal covariance)
  \item \textbf{VEV:} varying volume, equal shape, varying orientation (ellipsoidal covariance)
  \item \textbf{VVV:} varying volume, varying shape, varying orientation (ellipsoidal covariance)
  \item \textbf{VEE:} Varying volume, Equal shape, Equal orientation.
  \item \textbf{EVE:} Equal volume, Varying shape, Equal orientation.
  \item \textbf{VVE:} Varying volume, Varying shape, Equal orientation.
  \item \textbf{EVV:} Equal volume, Varying shape, Varying orientation.
\end{itemize}
```
Interestingly, this outcome differs from a previous analysis where the
best number of clusters, determined through computing the ELBO, was
found to be $K=2$. Nevertheless, recent studies Zhang2024 suggest that
ELBO tends to achieve better approximation accuracy than BIC with
similar computational costs when estimating model evidence. This result
is reasonable, considering that from a visual representation of the
data, they appear to cluster around two distinct points.

# References

Zhang, Y., & Yang, Y. (2024). "Bayesian model selection via mean-field
variational approximation." Journal of the Royal Statistical Society
Series B: Statistical Methodology. doi: 10.1093/jrsssb/qkad164.

\clearpage

# Appendix

# Targetscore package
```{r}
# Variational Bayesian Gaussian mixture model (VB-GMM)
# X: N x D data matrix
# init: k (1 x 1) or label (1 x n, 1<=label(i)<=k) or center (d x k)
# Reference: Pattern Recognition and Machine Learning by Christopher M. Bishop (P.474)
# Part of the implementation is based on the Matlab code vbgm.m from Michael Chen
# Matlab code: http://www.mathworks.com/matlabcentral/fileexchange/35362-variational-bayesian-inference-for-gaussian-mixture-model
# Author: Yue Li (yueli@cs.toronto.edu)

# Main function vbgmm running four separate functions, namely,
# intialization: initialization of responsibility
# vexp: Variational-Expectation
# vmax: V-Maximimization
# vbound: V-(lower)-bound evaluation
require(pracma)
require(Matrix)


############ bsxfun {pracma} with single expansion (real Matlab style) ############
# expandByRow applies only when x is a square matrix
bsxfun.se <- function(func, x, y, expandByRow=TRUE) {
	
	if(length(y) == 1) return(arrayfun(func, x, y)) else	
		stopifnot(nrow(x) == length(y) || ncol(x) == length(y))
	
	expandCol <- nrow(x) == length(y)	
	expandRow <- ncol(x) == length(y)			
	
	if(expandCol & expandRow & expandByRow) expandCol <- FALSE
	if(expandCol & expandRow & !expandByRow) expandRow <- FALSE
			
	# repeat row (if dim2expand = 1, then length(y) = ncol(x))
	if(expandRow) y.repmat <- matrix(rep(as.numeric(y), each=nrow(x)), nrow=nrow(x))
	
	# repeat col (if dim2expand = 2, then length(y) = nrow(x))
	if(expandCol) y.repmat <- matrix(rep(as.numeric(y), ncol(x)), ncol=ncol(x))
	
	bsxfun(func, x, y.repmat)	
}


############ Compute log(sum(exp(x), dim)) ############
# Compute log(sum(exp(x),dim)) while avoiding numerical underflow
logsumexp <- function(x, margin=1) {
	
	stopifnot(is.matrix(x))
	
	# subtract the largest in each column
	y <- apply(x, margin, max)
					
	x <- bsxfun.se("-", x, y)
	
	s <- y + log(apply(exp(x), margin, sum))
	
	i <- which(!is.finite(s))
	
	if(length(i) > 0) s[i] <- y[i]
	
	s
}


############ Logarithmic Multivariate Gamma function ############
# Compute logarithm multivariate Gamma function.
# Gamma_p(x) = pi^(p(p-1)/4) prod_(j=1)^p Gamma(x+(1-j)/2)
# log Gamma_p(x) = p(p-1)/4 log pi + sum_(j=1)^p log Gamma(x+(1-j)/2)
logmvgamma <- function(x, d) {
	
	s <- size(x)
	
	x <- matrix(as.numeric(x), nrow=1)
			
	x <- bsxfun.se("+", kronecker(matrix(1,d,1), x), (1 - matrix(1:d))/2)
	
	y <- d*(d-1)/4*log(pi) + colSums(lgamma(x))
	
	y <- matrix(as.numeric(y), nrow=s[1], ncol=s[2])
	
	y
}

############ Sort the model  ############
# Sort model paramters in increasing order of averaged means
# of d variables
sort_components <- function(model) {
	
	idx <- order(apply(model$m, 2, mean))
	
	model$m <- model$m[, idx]
	
	model$R <- model$R[, idx]
	
	model$logR <- model$logR[, idx]
		
	model$alpha <- model$alpha[idx]
	
	model$kappa <- model$kappa[idx]
	
	model$v <- model$v[idx]
	
	model$M <- model$M[,,idx]
	
	model
}


############ Vector dot product ############
# handle single row matrix by multiplying each value
# but not sum them up
dot.ext <- function(x,y,mydim) {
	
	if(missing(mydim)) dot(x,y) else {
				
		if(1 %in% size(x) & mydim == 1) x * y else dot(x,y)		
	}
}



############ Main function of VB-GMM ############
# mirprior: miRNA-specific prior

vbgmm <- function(data, init=2, prior, tol=1e-20, maxiter=2e3, mirprior=TRUE, 
	expectedTargetFreq=0.01, verbose=FALSE) {
	
	data <- as.matrix(data)
	
	n <- nrow(data)
	d <- ncol(data)
	
	X <- t(data) # Work with D by N for convenience		
	
	message(sprintf("Running VB-GMM on a %d-by-%d data ...\n", n, d))
	
	if(missing(prior)) {
		
		# miRNA-specific prior with alpha for target-component
		# with defined target frequency (expected much lower
		# than non-target freq)
		if(mirprior & length(init) == 1) {
			
			k <- init
			
			stopifnot(expectedTargetFreq > 0 & expectedTargetFreq < 1)
									
			expectTargetCnt <- round(expectedTargetFreq * n)
			
			backgroundCnt <- rep((n - expectTargetCnt)/(k-1), k-1)
			
			prior <- list(
				alpha = c(expectTargetCnt, backgroundCnt),
				kappa = 1,
				m = as.matrix(rowMeans(X)),
				v = d+1,
				M = diag(1,d,d) # M = inv(W)
			)
			
		} else { # more general prior with equal alpha
			
			if(length(init)>1) k <- ncol(init) else k <- init
			
			prior <- list(
				alpha = rep(1,k),
				kappa = 1,
				m = as.matrix(rowMeans(X)),
				v = d+1,
				M = diag(1,d,d) # M = inv(W)
			)
		}
	} else {
		
		stopifnot(
			all(names(prior) %in% c("alpha","kappa","m","v","M")) &
			all(sapply(prior, is.numeric)) & nrow(prior$m) == d &
			ncol(prior$m) == 1 &
			nrow(prior$M) == d & ncol(prior$M) == d)					
	}
	
	if(mirprior & length(init) != 1) {
		warning("mirprior is TRUE but init is not scalar (k)! Proceed as if mirprior were FALSE") }
	
	# lower variational bound (objective function)
	L <- rep(-Inf, maxiter)	
	converged <- FALSE	
	t <- 1
	
	model <- list()
	
	model$R <- initialization(X, init) # initialize responsibility R
	
	while(!converged & t < maxiter) {				
		t <- t + 1
		model <- vmax(X, model, prior)
		model <- vexp(X, model)
		L[t] <- vbound(X, model, prior)/n
		converged <- abs(L[t] - L[t-1]) < tol * abs(L[t])
		
		if(verbose) message(sprintf("VB-EM-%d: L = %.6f", t, L[t]))
	}
	
	L <- L[2:t]
	
	model <- sort_components(model)
	
	label <- rep(0, n)
	
	label <- apply(model$R, 1, which.max)		
	
	# unique to have consecutive label eg 2,3,6 changed to 1,2,3
	# label <- match(label, sort(unique(label)))	
	
	if(converged) message(sprintf("Converged in %d steps.\n", t-1)) else
	warnings(sprintf("Not converged in %d steps.\n", maxiter))
	
	list(label=label, R=model$R, mu=model$m, full.model=model, L=L)
}


############ Initialization of responsibility (intialization) ############
initialization <- function(X, init) {
	
	d <- nrow(X)
		
	n <- ncol(X)
	
	stopifnot(length(init) %in% c(1, n) ||
		(nrow(init) == d  & ncol(init) == k))
		
	if(length(init) == 1) { # init = k gaussian components
		
		k <- init
		
		idx <- sample(1:n, k)
		
		m <- X[,idx,drop=F]		
		
		label <- apply(bsxfun.se("-", crossprod(m, X),
			as.matrix(dot.ext(m,m,1)/2)), 2, which.max)
		
		# unique to have consecutive label eg 2,3,6 changed to 1,2,3
		label <- match(label, sort(unique(label)))
		
		while(k != length(unique(label))) {
			
			idx <- sample(1:n, k)
		
			m <- X[,idx,drop=F]
			
			label <- apply(bsxfun.se("-", crossprod(m, X),
				as.matrix(dot.ext(m,m,1)/2)), 2, which.max)
				
			label <- match(label, unique(label))			
		}
		
		R <- as.matrix(sparseMatrix(1:n, label, x=1))
		
	} else {
	
		if(length(init) == n) { # initialize with labels
			
			label <- init		
			k <- max(label)		
			R <- as.matrix(sparseMatrix(1:n, label, x=1))
			
		} else {
		
			if(!is.null(dim(init))) {
				
				if(nrow(init) == d  & ncol(init) == k) { # initialize with centers
				
					k <- ncol(init)		
					m <- init		
					label <- apply(bsxfun.se("-", crossprod(m, X),
						as.matrix(dot.ext(m,m,1)/2)), 2, which.max)		
					R <- as.matrix(sparseMatrix(1:n, label, x=1))
					
				} else stop(message("Invalid init."))				
			}
		}
	}

	R	
}


############ Variational-Maximimization ############
vmax <- function(X, model, prior) {
					
	alpha0 <- prior$alpha	
	kappa0 <- prior$kappa
	m0 <- prior$m;
	v0 <- prior$v;
	M0 <- prior$M;
	R <- model$R;

	nk <- apply(R, 2, sum) # 10.51
	alpha <- alpha0 + nk # 10.58
	nxbar <- X %*% R
	kappa <- kappa0 + nk # 10.60		
	m <- bsxfun.se("*", bsxfun.se("+", nxbar, kappa0 * m0), 1/kappa) # 10.61
	v <- v0 + nk # 10.63 (NB: no 1 in the matlab code)
	
	d <- nrow(m)
	k <- ncol(m)
	
	M <- array(0, c(d, d, k))
	sqrtR <- sqrt(R)
	
	xbar <- bsxfun.se("*", nxbar, 1/nk) # 10.52
	xbarm0 <- bsxfun.se("-", xbar, m0)
	w <- (kappa0 * nk) * (1/(kappa0 + nk))
	
	for(i in 1:k) {
		
		Xs <- bsxfun.se("*", bsxfun.se("-", X, xbar[,i]), t(sqrtR[,i]))
		
		xbarm0i <- xbarm0[,i]
		
		# 10.62
		M[,,i] <- M0 + tcrossprod(Xs, Xs) + w[i] * tcrossprod(xbarm0i, xbarm0i)
	}		
	
	model$alpha <- alpha
	model$kappa <- kappa
	model$m <- m
	model$v <- v
	model$M <- M # Whishart: M = inv(W)
	
	model
}


############ Variational-Expectation ############
vexp <- function(X, model) {
	
	alpha <- model$alpha	# Dirichlet
	kappa <- model$kappa	# Gaussian
	m <- model$m			# Gasusian
	v <- model$v			# Whishart
	M <- model$M			# Whishart: inv(W) = V'*V
	
	n <- ncol(X)
	d <- nrow(m)
	k <- ncol(m)
	
	logW <- array(0, dim=c(1,k))
	EQ <- array(0, dim=c(n,k))
	
	for(i in 1:k) {
		
		U <- chol(M[,,i])
		
		logW[i] <- -2 * sum(log(diag(U)))
		
		Q <- solve(t(U), bsxfun.se("-", X, m[,i]))
		
		EQ[,i] <- d/kappa[i] + v[i] * dot.ext(Q,Q,1)	# 10.64
	}
	
	vd <- bsxfun.se("-", matrix(rep(v+1, d),nrow=d,byrow=T), as.matrix(1:d))/2
	
	ElogLambda <- colSums(digamma(vd)) + d*log(2) + logW	# 10.65	
	Elogpi <- digamma(alpha) - digamma(sum(alpha))			# 10.66
	
	logRho <- (bsxfun.se("-", EQ, 2*Elogpi + ElogLambda - d*log(2*pi)))/(-2) # 10.46
	logR <- bsxfun.se("-", logRho, logsumexp(logRho, 1))	# 10.49
	R <- exp(logR)
	
	model$logR <- logR
	model$R <- R
	
	model
}



############ Variational-(lower)-Bound Evaluation ############
vbound <- function(X, model, prior) {
	
	alpha0 <- prior$alpha
	kappa0 <- prior$kappa
	m0 <- prior$m
	v0 <- prior$v
	M0 <- prior$M

	alpha <- model$alpha	# Dirichlet
	kappa <- model$kappa	# Gaussian
	m <- model$m			# Gasusian
	v <- model$v			# Whishart
	M <- model$M			# Whishart: inv(W) = V'*V
	R <- model$R
	logR <- model$logR
	
	d <- nrow(m)
	k <- ncol(m)
	nk <- colSums(R)									# 10.51
	
	Elogpi <- digamma(alpha) - digamma(sum(alpha))		# 10.66
	
	Epz = dot(nk, Elogpi)								# 10.72
	Eqz = dot(as.numeric(R), as.numeric(logR))			# 10.75
	# logCalpha0 = lgamma(k * alpha0) - k * lgamma(alpha0) # for scalar alpha0
	logCalpha0 = lgamma(sum(alpha0)) - sum(lgamma(alpha0))
	# Eppi <- logCalpha0+(alpha0-1)*sum(Elogpi) # for scalar alpha0
	Eppi <- logCalpha0+dot(alpha0-1, Elogpi)			# 10.73		
	logCalpha <- lgamma(sum(alpha))-sum(lgamma(alpha))
	Eqpi = dot(alpha-1, Elogpi) + logCalpha				# 10.76
	
	# part of 10.70
	L <- Epz - Eqz + Eppi - Eqpi
	
	U0 <- chol(M0)
	sqrtR <- sqrt(R)
	xbar <- bsxfun.se("*", X %*% R, 1/nk)				# 10.52
	
	logW <- array(0, dim = c(1, k))
	trSW <- array(0, dim = c(1, k))
	trM0W <- array(0, dim = c(1, k))
	xbarmWxbarm <- array(0, dim = c(1, k))
	mm0Wmm0 <- array(0, dim = c(1, k))
	
	for(i in 1:k) {
		
		U <- chol(M[,,i])
		logW[i] <- -2 * sum(log(diag(U)))
		
		Xs <- bsxfun.se("*", bsxfun.se("-", X, as.matrix(xbar[,i,drop=F])), t(sqrtR[,i,drop=F]))
		V <- chol(tcrossprod(Xs, Xs)/nk[i])
		Q <- solve(U, V)
		# equivalent to tr(SW)=trace(S/M)
		trSW[i] <- dot(as.numeric(Q), as.numeric(Q))
		Q <- solve(U, U0)
		trM0W[i] <- dot(as.numeric(Q), as.numeric(Q))
		
		q <- solve(t(U), xbar[,i,drop=F]-m[,i,drop=F])
		xbarmWxbarm[i] = dot(q, q)
		q <- solve(t(U), m[,i,drop=F]-m0)
		mm0Wmm0[i] <- dot(q, q)
	}
	
	vd <- bsxfun.se("-", matrix(rep(v+1, d),nrow=d,byrow=T), as.matrix(1:d))/2	
	ElogLambda <- colSums(digamma(vd)) + d*log(2) + logW	# 10.65
	
	# first half of 10.74
	Epmu <- sum(d*log(kappa0/(2*pi))+ElogLambda-d*kappa0/kappa-kappa0*(v*mm0Wmm0))/2
	logB0 <- v0*sum(log(diag(U0)))-0.5*v0*d*log(2)-logmvgamma(0.5*v0,d) # B.79
	# second half of 10.74	
	EpLambda <- k*logB0+0.5*(v0-d-1)*sum(ElogLambda)-0.5*dot(v,trM0W)
	
	
	Eqmu <- 0.5*sum(ElogLambda+d*log(kappa/(2*pi)))-0.5*d*k	# 10.77 (1/2)
	logB <- (-v) * (logW+d*log(2))/2 - logmvgamma(0.5*v, d)	# B.79
	EqLambda <- 0.5*sum((v-d-1)*ElogLambda-v*d)+sum(logB) 	# 10.77 (2/2)
	
	EpX <- 0.5*dot(nk, ElogLambda-d/kappa-v*trSW-v*xbarmWxbarm-d*log(2*pi))	# 10.71
		
	L <- L+Epmu-Eqmu+EpLambda-EqLambda+EpX	# 10.70
	
	L
}
```

